---
title: "Selection Bias in Vaccine Data"
author: "David Casente"
date: "4/27/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Notes



Odds of a state being censored: 4 / 51

L is State
Y is Vaccination




```{r import, echo=FALSE}
library(readr)
library(dplyr)
library(ggplot2)

vacc <- read_csv("vacc.csv", 
    col_types = cols(`% of Vaccinations with Known Ethnicity` = col_double(), 
        `% of Vaccinations with Known Race` = col_double(), 
        `% of Vaccinations with Unknown Ethnicity` = col_double(), 
        `% of Vaccinations with Unknown Race` = col_double(), 
        `American Indian or Alaska Native % of Vaccinations` = col_double(), 
        `asian_vacc` = col_double(), 
        `black_vacc` = col_double(), 
        `hispanic_vacc` = col_double(), 
        `Native Hawaiian or Other Pacific Islander % of Vaccinations` = col_double(), 
        `Other % of Vaccinations` = col_double(), 
        `include_hispanic` = col_character(), 
        `white_vacc` = col_double()))

pop_race <- read_csv("state_pop_by_race.csv")
pop <- read_csv("state_population.csv")

```

## General Idea behind the IP weighting



You can also embed plots, for example:

```{r, echo=FALSE}
vacc_clean <- vacc %>% 
  select(location, white_vacc, black_vacc, hispanic_vacc, asian_vacc) %>%
  mutate(including_hispanic = ifelse(is.na(hispanic_vacc), 0,1))



pop_race_clean <- pop_race %>%
  select(location, white_pop, black_pop, hispanic_pop, asian_pop )


pop_clean <- pop %>%
  select(location, "2019") %>%
  filter(location != "Puerto Rico" & location != "Northeast" & location != "United States" & location != "Midwest" & location != "South" & location != "West")

```



let's plot out all the population and treatment variables to see the grouping of the data and how to best bin them

```{r,echo=FALSE}
for (i in colnames(pop_race_clean)){
  if(class(pop_race_clean[[i]]) == "numeric"){
    temp <- density(pop_race_clean[[i]])
    
    plot(temp, main = paste("Density plot of", i))
  }
}


pop_dplot <- density(pop_clean$`2019`)
plot(pop_dplot, main = "Density plot of population")




```




Need to make seperate dataframes for uncensored and censored for each statistic we are filling out: black_pop, asian_pop, white_pop, hispanic_pop

```{r}

master_df <- merge(pop_clean, pop_race_clean, by = "location")

master_df <- merge(master_df, vacc_clean, by = "location")


master_df_clean <- master_df %>%
  select(!including_hispanic) %>%
  rename(state_pop = "2019")

white <- master_df_clean %>% 
  select(location, state_pop, white_pop, white_vacc)

white_uc <- white %>%
  filter(!is.na(white_vacc))


white_c <- white %>%
  filter(is.na(white_vacc))

black <- master_df_clean %>% 
  select(location, state_pop, black_pop, black_vacc)


black_uc <- black %>%
  filter(!is.na(black_vacc))


black_c <- black %>%
  filter(is.na(black_vacc))

hispanic <- master_df_clean %>% 
  select(location, state_pop, hispanic_pop, hispanic_vacc)
  
hispanic_uc <- hispanic %>%
  filter(!is.na(hispanic_vacc))


hispanic_c <- hispanic %>%
  filter(is.na(hispanic_vacc))


asian <- master_df_clean %>% 
  select(location, state_pop, asian_pop, asian_vacc)

asian_uc <- asian %>%
  filter(!is.na(asian_vacc))


asian_c <- asian %>%
  filter(is.na(asian_vacc))
  





```



define bins for ipw tree

```{r}
pop_bins <- c(0,3000000,10000000,50000000)
white_bins <- c(0,0.55,0.65,0.75, 0.85, 1)
black_bins <- c(0,0.05,0.1,0.15,0.2,1)
hispanic_bins <- c(0,0.075,0.125,0.175,0.225,1)
asian_bins <- c(0,0.02,0.05,1)

```


Now we can loop throught for each censorec state and find the correct value through ipw






```{r}
fill_in_the_blanks_white <- function(censored,uncensored, pop_state_bins, race_bins) {
  
  for(j in 1:nrow(censored)){
    state_num <- 2
    race_num <- 3
    vacc_num <- 4
    
    lower_state <- get_pop_state_lower(censored[j,state_num],pop_state_bins)
    upper_state <- get_pop_state_upper(censored[j,state_num],pop_state_bins)
    lower_race <- get_pop_race_lower(censored[j,race_num],race_bins)
    upper_race <- get_pop_race_upper(censored[j,race_num],race_bins)
    
    
    vacc_pop <- colnames(uncensored)[vacc_num]
    
    print(paste("location: ", censored[j,1]," ls: ", lower_state, " us: ", upper_state," lr: ", lower_race," ur: ", upper_race))
    
    uc_group <- uncensored %>% 
      filter(state_pop > lower_state & state_pop < upper_state) %>%
      filter(white_pop > lower_race & white_pop < upper_race)
    
    print(uc_group)
    
    censored[j,vacc_num] <- mean(uc_group[["white_vacc"]])
    
    
    
  }
  
  return(censored)
}


fill_in_the_blanks_black <- function(censored,uncensored, pop_state_bins, race_bins) {
  
  for(j in 1:nrow(censored)){
    state_num <- 2
    race_num <- 3
    vacc_num <- 4
    
    lower_state <- get_pop_state_lower(censored[j,state_num],pop_state_bins)
    upper_state <- get_pop_state_upper(censored[j,state_num],pop_state_bins)
    lower_race <- get_pop_race_lower(censored[j,race_num],race_bins)
    upper_race <- get_pop_race_upper(censored[j,race_num],race_bins)
    
    
    vacc_pop <- colnames(uncensored)[vacc_num]
    
    print(paste("location: ", censored[j,1]," ls: ", lower_state, " us: ", upper_state," lr: ", lower_race," ur: ", upper_race))
    
    uc_group <- uncensored %>% 
      filter(state_pop > lower_state & state_pop < upper_state) %>%
      filter(black_pop > lower_race & black_pop < upper_race)
    
    print(uc_group)
    
    censored[j,vacc_num] <- mean(uc_group[["black_vacc"]])
    
    
    
  }
  
  return(censored)
}

fill_in_the_blanks_asian <- function(censored,uncensored, pop_state_bins, race_bins) {
  
  for(j in 1:nrow(censored)){
    state_num <- 2
    race_num <- 3
    vacc_num <- 4
    
    lower_state <- get_pop_state_lower(censored[j,state_num],pop_state_bins)
    upper_state <- get_pop_state_upper(censored[j,state_num],pop_state_bins)
    lower_race <- get_pop_race_lower(censored[j,race_num],race_bins)
    upper_race <- get_pop_race_upper(censored[j,race_num],race_bins)
    
    
    vacc_pop <- colnames(uncensored)[vacc_num]
    
    print(paste("location: ", censored[j,1]," ls: ", lower_state, " us: ", upper_state," lr: ", lower_race," ur: ", upper_race))
    
    uc_group <- uncensored %>% 
      filter(state_pop > lower_state & state_pop < upper_state) %>%
      filter(asian_pop > lower_race & asian_pop < upper_race)
    
    print(uc_group)
    
    censored[j,vacc_num] <- mean(uc_group[["asian_vacc"]])
    
    
    
  }
  
  return(censored)
}

fill_in_the_blanks_hispanic <- function(censored,uncensored, pop_state_bins, race_bins) {
  
  for(j in 1:nrow(censored)){
    state_num <- 2
    race_num <- 3
    vacc_num <- 4
    
    lower_state <- get_pop_state_lower(censored[j,state_num],pop_state_bins)
    upper_state <- get_pop_state_upper(censored[j,state_num],pop_state_bins)
    lower_race <- get_pop_race_lower(censored[j,race_num],race_bins)
    upper_race <- get_pop_race_upper(censored[j,race_num],race_bins)
    
    
    vacc_pop <- colnames(uncensored)[vacc_num]
    
    print(paste("location: ", censored[j,1]," ls: ", lower_state, " us: ", upper_state," lr: ", lower_race," ur: ", upper_race))
    
    uc_group <- uncensored %>% 
      filter(state_pop > lower_state & state_pop < upper_state) %>%
      filter(hispanic_pop > lower_race & hispanic_pop < upper_race)
    
    print(uc_group)
    
    censored[j,vacc_num] <- mean(uc_group[["hispanic_vacc"]])
    
    
    
  }
  
  return(censored)
}


white_filled_in <- fill_in_the_blanks_white(white_c,white_uc,pop_bins,white_bins)
black_filled_in <- fill_in_the_blanks_black(black_c,black_uc,pop_bins,black_bins)
asian_filled_in <- fill_in_the_blanks_asian(asian_c,asian_uc,pop_bins,asian_bins)
hispanic_filled_in <- fill_in_the_blanks_hispanic(hispanic_c,hispanic_uc,pop_bins,hispanic_bins)



```


combine censored and uncensored
```{r}
white_combo <- rbind(white_uc, white_filled_in)
hispanic_combo <- rbind(hispanic_uc, hispanic_filled_in)
black_combo <- rbind(black_uc, black_filled_in)
asian_combo <- rbind(asian_uc, asian_filled_in)



output <- merge(white_combo,hispanic_combo, by = c("location","state_pop"))
output <- merge(output, black_combo, by = c("location","state_pop"))
output <- merge(output, asian_combo, by = c("location","state_pop"))

output <- output %>%
  select(location, state_pop, white_pop, black_pop, hispanic_pop, asian_pop, white_vacc, black_vacc, hispanic_vacc, asian_vacc)

write_csv(output, "filled_in_data.csv")



```

functions to get right boundaries
```{r}
get_pop_state_upper <- function(state_pop_c, bins) {
  upper_bound <- 1
  
  for(i in 1:length(bins)){
    if(state_pop_c < bins[length(bins) + 1 - i]){
      upper_bound <- bins[length(bins) + 1 - i]
      
    }
  }

  return(upper_bound)
}


get_pop_state_lower <- function(state_pop_c, bins) {
  lower_bound <- 0
  
  for(i in 1:length(bins)){
    if(state_pop_c > bins[i]){
      lower_bound <- bins[i]
    }
  }
  
  return(lower_bound)
}


get_pop_race_upper <- function(state_race_c, bins) {
  upper_bound <- 1
  
  for(i in 1:length(bins)){
    if(state_race_c < bins[length(bins) + 1 - i]){
      upper_bound <- bins[length(bins) + 1 - i]
      
    }
  }
  
  
  return(upper_bound)
}


get_pop_race_lower <- function(state_race_c, bins) {
  lower_bound <- 0
  
  for(i in 1:length(bins)){
    if(state_race_c > bins[i]){
      lower_bound <- bins[i]
    }
  }
  
  
  return(lower_bound)
}










```
















graveyard









```{r}
fill_in_the_blanks <- function(censored,uncensored, pop_state_bins, race_bins, state_column_name, race_column_name, vacc_column_name) {
  
  for(j in 1:length(censored)){
    state_num <- 2
    race_num <- 3
    vacc_num <- 4
    
    lower_state <- get_pop_state_lower(censored[j,state_num],pop_state_bins)
    upper_state <- get_pop_state_upper(censored[j,state_num],pop_state_bins)
    lower_race <- get_pop_race_lower(censored[j,race_num],race_bins)
    upper_race <- get_pop_race_upper(censored[j,race_num],race_bins)
    
    
    vacc_pop <- colnames(uncensored)[vacc_num]
    
    print(paste("location: ", censored[j,1]," ls: ", lower_state, " us: ", upper_state," lr: ", lower_race," ur: ", upper_race))
    
    uc_group <- uncensored %>% 
      filter(state_column_name > lower_state & state_column_name < upper_state) %>%
      filter(race_column_name > lower_race & race_column_name < upper_race)
    
    print(uc_group)
    
    censored[j,vacc_num] <- mean(uc_group[[vacc_column_name]])
    
    
    
  }
  
  return(censored)
}

```